---
title: "Lab 08 - Text Mining/NLP"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(eval = T, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r}
mt_samples |>
  count(medical_specialty, sort = TRUE) |>
  ggplot(aes(fct_reorder(medical_specialty, n), n))+
  geom_col(fill="dodgerblue")+
  coord_flip()+
  theme_bw()
```
The bar plot shows that some medical specialties have many more records than others. Surgery, Orthopedic, and Radiology appear the most, while Dentistry, Cardiovascular/Pulmonary, and Allergy/Immunology have very few. Some specialties may have similar topics, like Neurosurgery and Neurology or Psychiatry and Pain Management. This uneven distribution might be due to how the data was collected. 

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarise(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
tokens
```

```{r}
tokens |>
  ggplot(aes(fct_reorder(word, word_frequency), word_frequency)) +
  geom_bar(stat = "identity", fill="dodgerblue") +
  coord_flip()+
  theme_bw()

tokens |>
  count(word, sort = TRUE) |>
  wordcloud2(size = 0.4, color = "random-light", backgroundColor = "dodgerblue")
```
The result shows common words like "the," "and," and "was," which do not give useful insights. The word "patient" appears often, suggesting the data is about medical records. Removing stopwords can help find important medical terms.
---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

```{r}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription, token = "words") |>
  anti_join(stop_words, by = "word") |> 
  filter(!str_detect(word, "^[0-9]+$")) |>  
  filter(!word %in% c("mm", "mg", "noted")) |>
  count(word, sort = TRUE) |>  
  top_n(20, n)  

tokens |>
  ggplot(aes(x = n, y = fct_reorder(word, n))) +
  geom_col(fill = "dodgerblue") +
  theme_bw()

tokens |>
  count(word, sort = TRUE) |>
  wordcloud2(size = 0.4, color = "random-light", backgroundColor = "dodgerblue")
```

After removing stopwords and filtering further, we see more meaningful medical terms. Words like "patient," "procedure," "history," "pain," "anesthesia," and "diagnosis" appear frequently, showing a focus on medical procedures and conditions. This helps us understand the dataset better, as these terms relate directly to patient care and treatments. Compared to the unfiltered data, this version gives a clearer picture of the main topics in the medical records.
---



## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r}
stop_words2<-c("mm","mg","noted",stop_words$word)
sw_start <- paste0("^", paste(stop_words2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stop_words2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!grepl(sw_start, ngram, ignore.case=TRUE))|>
  filter(!grepl(sw_end, ngram, ignore.case = TRUE))|>
  filter(!grepl("[[:digit:]]+", ngram))|>
  group_by(ngram)|>
  summarise(word_frequency=n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
```

```{r}

tokens_bigram |>
  ggplot(aes(ngram, word_frequency)) +  
  geom_col(fill = "dodgerblue") + 
  coord_flip() + 
  theme_bw()

```
Bi-grams show more meaningful medical phrases like "preoperative diagnosis" and "blood pressure," giving better context than single words. Tri-grams would be even more specific, like "history of present illness," but may appear less often. Removing stopwords in n-grams needs extra care to keep useful medical terms.
---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r}
library(stringr)
# e.g. patient, blood, preoperative...
tokens_bigram |>
  filter(str_detect(ngram, regex("\\sblood$|^blood\\s"))) |>
  mutate(word = str_remove(ngram, "blood"),
  word = str_remove_all(word, " ")) |>
  group_by(word) |>
  head(20)|>
  ggplot(aes(reorder(word, word_frequency),word_frequency)) +geom_col(fill="dodgerblue") +theme_bw()
  
```

---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r}

mt_samples |>
   unnest_tokens(word, transcription) |>
   anti_join(stop_words, by = "word") |>  
   filter(!str_detect(word, "^[0-9]+$")) |> 
   count(medical_specialty, word, sort = TRUE) |>  
   group_by(medical_specialty) |> 
   slice_max(order_by = n, n = 5) |>  
   ungroup()

mt_samples |>
   unnest_tokens(word, transcription) |>
   anti_join(stop_words, by = "word") |>  
   filter(!str_detect(word, "^[0-9]+$")) |>  
   count(word, sort = TRUE) |>  
   slice_max(order_by = n, n = 5)

```
The table shows the most used words in each medical specialty, and the overall top five is patient, left that might realted to "left arm" or "left leg" and procedure, history,normal, which is common word used in hospitcal related feild.

## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r}

transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by = "word") |>
  filter(!str_detect(word, "^[0-9]+$")) |>  
  filter(!word %in% c("mm", "mg", "noted")) |>  
  DocumentTermMatrix()

```

```{r}
library(reshape2)

transcripts_dtm <- as.matrix(transcripts_dtm)   

transcripts_lda <- LDA(transcripts_dtm, k = 3, 
  control = list(seed = 1234))


transcripts_top_terms <-
  tidy(transcripts_lda, matrix="beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |>
  group_by(topic) |>
  slice_max(beta, n=10) |>
  ungroup() |>
  arrange(topic, -beta)

transcripts_top_terms |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales="free") +
  scale_y_reordered() +
  theme_bw()

```




